---
title: "Interactive Visualizations"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

```{r setup, message=FALSE, echo=FALSE, error=FALSE, warning=FALSE}
library(httr)
library(jsonlite)
library(dplyr)
library(glue)
library(stringr)
library(tibble)
library(opendatatoronto)
library(knitr)
library(kableExtra)
library(ggplot2)
library(ggcorrplot)
library(wordcloud)
library(RColorBrewer)
library(tidyr)
library(rpart)
library(rpart.plot)
library(randomForest)
library(gbm)
library(xgboost)
library(caret)
library(mgcv)
library(plotly)
library(webshot2)
library(leaflet)
```


```{r, message=FALSE, echo=FALSE, error=FALSE, warning=FALSE}
hd2021_census <- read.csv("Data_Wrangling/hd2021_census.csv", check.names = FALSE)
census_data <- read.csv("data/census_data.csv", check.names = FALSE)
auto_theft_census <- read.csv("data/auto_theft_census.csv", check.names = FALSE )
```

```{r, message=FALSE, echo=FALSE, error=FALSE, warning=FALSE}
set.seed(1250)
train <- sample(1:nrow(auto_theft_census), round(0.7 * nrow(auto_theft_census)))
train_data <- auto_theft_census[train, ]
test_data <- auto_theft_census[-train, ]
```


<br>

## Auto Theft Neighbourhood Density
Leaflet map that visualizes the spatial distribution of auto theft incidents across different neighbourhoods in Toronto, providing an interactive way to examine crime density and making it easier to quickly identify areas that may require targeted interventions or additional resources.
```{r figure1, message=FALSE, echo=FALSE, warning=FALSE}
neighbourhood_summary <- train_data |>
  filter(!is.na(NEIGHBOURHOOD_140)) |>
  group_by(NEIGHBOURHOOD_140) |>
  summarize(Total_Auto_Theft = n(), .groups = "drop")

neighbourhood_locations <- train_data |>
  filter(!is.na(NEIGHBOURHOOD_140) & !is.na(LAT_WGS84) & !is.na(LONG_WGS84)) |>
  group_by(NEIGHBOURHOOD_140) |>
  summarize(lat = mean(LAT_WGS84), lon = mean(LONG_WGS84), .groups = "drop")

crime_map_data <- left_join(neighbourhood_summary, neighbourhood_locations, by = "NEIGHBOURHOOD_140")

pal = colorNumeric(palette = rev(viridis::magma(256)), domain = crime_map_data$Total_Auto_Theft)

leaflet(crime_map_data) |>
  addTiles() |>  
  addCircleMarkers(
    ~lon, ~lat, 
    radius = ~sqrt(Total_Auto_Theft) * 2, 
    color = ~pal(Total_Auto_Theft),
    fillColor = ~pal(Total_Auto_Theft),
    fillOpacity = 1.0,
    popup = ~paste0("<b>neighbourhood: </b>", NEIGHBOURHOOD_140, "<br>",
                    "<b>Total Crimes: </b>", Total_Auto_Theft)
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = crime_map_data$Total_Auto_Theft,
    title = "Crime Count",
    labels = "Higher crimes → Bigger circles"
  ) |>
  addControl(
    html = "<h4 style='color:black; font-weight: bold;'>Auto Theft and Socio-Economic Analysis in Toronto Neighbourhoods</h4>",
    position = "bottomleft"
  )
```

The data for this interaction is a combination of number auto thefts from 2015 added with the number of auto thefts from 2020, so it shows the **total** number of auto thefts in 2015 and 2020 across all 140 Toronto neighbourhoods. The large black circle represents the area with the most total auto thefts in 2015 and 2020, which is West Humber-Clairville with 463 reported auto thefts. The reddish purplish points represent neighbourhoods on the higher end of the spectrum in terms of number of auto theft count, and these include York University Heights (200) and Islington-City Centre West (183). On the other hand, the smaller, light yellow circles represent the neighbourhoods with the least amount of auto thefts. From the plot, most areas appear to be under 150, West Humber-Clairville is considered an outlier.

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
train_data <- train_data |>
  group_by(REPORT_YEAR, `Neighbourhood Name`) |>
  mutate(Total_Auto_Thefts = n()) |>
  ungroup()  
```


```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
top_neighbourhoods <- train_data |>
  group_by(`Neighbourhood Name`) |>
  summarise(total_incidents = sum(Total_Auto_Thefts, na.rm = TRUE)) |>
  arrange(desc(total_incidents)) |>
  head(20) |>
  pull(`Neighbourhood Name`)

top_20_neighbourhoods <- train_data |>
  filter(`Neighbourhood Name` %in% top_neighbourhoods)

top_20_summary <- top_20_neighbourhoods |>
  group_by(`Neighbourhood Name`, REPORT_YEAR) |>
  summarise(Incidents = n(), .groups = "drop")
```

## Top 20 Neighbourhoods
This interactive double bar graph displays the number of auto thefts of the top 20 neighbourhoods with the most reported cases of auto theft, comparing numbers from both 2015 and 2020.
```{r figure2, message=FALSE, echo=FALSE, warning=FALSE}
plot_ly(
  data = top_20_summary,
  x = ~Incidents,
  y = ~`Neighbourhood Name`,
  color = ~factor(REPORT_YEAR),
  colors = c("lightblue", "lightgreen"),
  type = "bar",
  orientation = "h"
) |>
  layout(
    title = list(
      text = paste0(
        "Top 20 Neighbourhoods with Highest Auto Theft Incidents",
        "<br><sub><i>Double bar chart showing top 20 neighbourhoods with the highest number of auto thefts</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "Number of Incidents"),
    yaxis = list(title = "Neighbourhood"),
    barmode = "group",
    legend = list(title = list(text = "Year")),
    margin = list(t = 100, l = 100), 
    annotations = list(
      list(
        text = "Double Bar Graph of the Top 20 Neighbourhoods",
        xref = "paper", yref = "paper",
        x = 0.5, y = -0.2, showarrow = FALSE,
        font = list(size = 10, style = "italic")
      )
    )
  )
```

We can see more clearly here which neighbourhoods have the highest number of auto thefts, and we can see visually how the numbers increased from 2015 to 2020, as the blue bars, which correspond with 2015, are much shorter than the green bars, which correspond with 2020. This illustrates how within 5 years, there was indeed an increase in number of auto thefts in most Toronto neighbourhoods.

## Income Extremes {.tabset}
The following plots illustrate the highest and lowest average individual incomes, including West Humber-Clairville (Toronto’s top auto theft hotspot in 2015 and 2020) for comparison, to explore income as a potential confounding factor

### Income Extremes in 2015
```{r figure3, message=FALSE, echo=FALSE, warning=FALSE}
income_extremes_2015 <- train_data |>
  filter(REPORT_YEAR == 2015) |>
  distinct(NEIGHBOURHOOD_140, .keep_all = TRUE) |>
  arrange(`Average total income`) |>
  slice_head(n = 5) |>
  mutate(Category = "Lowest") |>  
  bind_rows(
    train_data |>
      filter(REPORT_YEAR == 2015) |>
      distinct(NEIGHBOURHOOD_140, .keep_all = TRUE) |>
      arrange(desc(`Average total income`)) |>
      slice_head(n = 5) |> 
      mutate(Category = "Highest") 
  ) |>
  bind_rows(
    train_data |>
      filter(REPORT_YEAR == 2015 & `Neighbourhood Name` == "West Humber-Clairville") |>
      distinct(NEIGHBOURHOOD_140, .keep_all = TRUE) |> 
      mutate(Category = "West Humber-Clairville") 
  )

options(scipen=999)

ggplot_income_plot <- ggplot(income_extremes_2015, aes(x = reorder(NEIGHBOURHOOD_140, -`Average total income`), 
                                                      y = `Average total income`, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Neighbourhoods with Highest & Lowest Average Income (2015)",
    x = "neighbourhood",
    y = "Average Household Income",
    fill = "Category"
  ) +
  scale_fill_manual(values = c("Highest" = "steelblue", "Lowest" = "red", "West Humber-Clairville" = "orange")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 11),  
    plot.caption = element_text(hjust = 0.5, face = "italic"),  
    legend.position = "bottom"
  ) +
  coord_flip()

ggplotly(ggplot_income_plot) |> 
  layout(
    annotations = list(
      list(
        x = 0.5,
        y = -0.2,
        text = "<i>Flipped bar plot depicting income extremes in 2015</i>",
        showarrow = FALSE,
        xref = "paper",
        yref = "paper",
        xanchor = "center",
        font = list(size = 10)
      )
    ),
    margin = list(b = 100)
  )
```

### Income Extremes in 2020
```{r figure4, message=FALSE, echo=FALSE, warning=FALSE}
income_extremes_2020 <- train_data |>
  filter(REPORT_YEAR == 2020) |>
  distinct(NEIGHBOURHOOD_140, .keep_all = TRUE) |>
  arrange(`Average total income`) |>
  slice_head(n = 5) |>
  mutate(Category = "Lowest") |>  
  bind_rows(
    train_data |>
      filter(REPORT_YEAR == 2020) |>
      distinct(NEIGHBOURHOOD_140, .keep_all = TRUE) |>
      arrange(desc(`Average after-tax income of household`)) |>
      slice_head(n = 5) |> 
      mutate(Category = "Highest")  
  ) |>
  bind_rows(
    train_data |>
      filter(REPORT_YEAR == 2020 & `Neighbourhood Name` == "West Humber-Clairville") |>
      distinct(NEIGHBOURHOOD_140, .keep_all = TRUE) |> 
      mutate(Category = "West Humber-Clairville")  
  )

ggplot_income_plot <- ggplot(income_extremes_2020, aes(x = reorder(NEIGHBOURHOOD_140, -`Average total income`), 
                                 y = `Average total income`, fill = Category)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Neighbourhoods with Highest & Lowest Average Income (2020)",
    x = "neighbourhood",
    y = "Average Household Income",
    fill = "Category",
    caption = "Flipped bar plot dipicting income extremes in 2020"
  ) +
  scale_fill_manual(values = c("Highest" = "steelblue", "Lowest" = "red", "West Humber-Clairville" = "orange")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 11),  
    plot.caption = element_text(hjust = 0.5, face = "italic"),  
    legend.position = "bottom"
) +
  coord_flip()

ggplotly(ggplot_income_plot) |> 
  layout(
    annotations = list(
      list(
        x = 0.5,
        y = -0.2,
        text = "<i>Flipped bar plot depicting income extremes in 2020</i>",
        showarrow = FALSE,
        xref = "paper",
        yref = "paper",
        xanchor = "center",
        font = list(size = 10)
      )
    ),
    margin = list(b = 100)
  )
```

## 
The plots above illustrate how West Humber-Clairville's average income aligns more closely with lower-income neighbourhoods in both 2015 and 2020. The plot also illustrates how many of the low income neighbourhoods, like Black Creek and Agincourt North, also rank high in auto thefts, as they are also seen in the `Top 20 Neighbourhoods with Highest Auto Theft Incidents` plot, suggesting a relationship between lower income and higher theft rates.


## Auto Thefts and Income Brackets
The follow presents a box plot grouping neighbourhoods into five income brackets to explore whether lower average individual income is associated with higher auto theft rates.
```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
options(scipen=999)

income_breaks <- c(0, 30000, 75000, 110000, 175000, 300000)
income_labels <- c(
  "<$30k", 
  "$30k-$75k", 
  "$75k-$110k", 
  "$110k-$170k", 
  "$170k-$300k"
)

train_data <- train_data |>
  mutate(Income_Bracket = cut(`Average total income`, 
                             breaks = income_breaks,
                             labels = income_labels,
                             include.lowest = TRUE))


plot_ly(
  data = train_data,
  x = ~Income_Bracket,
  y = ~Total_Auto_Thefts,
  color = ~factor(REPORT_YEAR),
  colors = c("lightpink", "lightgreen"),
  type = "box",
  boxpoints = "outliers"
) |>
  layout(
    title = "Auto Theft Incidents Across Income Groups",
    xaxis = list(
      title = "Income Brackets",
      tickformat = ",d",
      tickangle = 45,
      separatethousands = TRUE,  
      exponentformat = "none",
      caption = "    "
    ),
    yaxis = list(
      title = "Total Auto Thefts",
      tickformat = ",d",
      separatethousands = TRUE,  
      exponentformat = "none"
    ),
    legend = list(title = list(text = "Year")),
    margin = list(b = 100), 
    annotations = list(
      list(
        text = "Boxplot using 5 breaks analyzing auto theft incidents across income groups",
        xref = "paper", yref = "paper",
        x = 0.5, y = -0.27, showarrow = FALSE,
        font = list(size = 10, style = "italic")
      )
    )
  )
```

This plot examines whether lower average individual income correlates with higher auto theft rates. It is important to note that in 2020, there are no neighbourhoods within the lowest income bracket (below $30K), likely reflecting broader income shifts over time. The figure illustrates that neighbourhoods in lower income brackets tend to experience greater variability in auto theft incidents, as indicated by wider interquartile ranges and longer tails. In contrast, higher-income brackets exhibit less variation, with narrower boxes and fewer extreme values. Notably, the highest recorded auto theft count, which was previously identified as West Humber-Clairville, falls within one of the lower income brackets (30K-75K) in both 2015 and 2020, once again reinforcing the relationship between lower neighbourhood income levels and higher auto theft rates.


## Age Distribution
This plot shows age group trends across the top 5 Toronto neighbourhoods with the most auto thefts, using percentage distributions of each age group by neighbourhood for both 2015 and 2020.

```{r echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
total_auto_theft <- train_data |>
  group_by(NEIGHBOURHOOD_140) |>
  summarise(Total_Auto_Theft = n()) |>
  arrange(desc(Total_Auto_Theft))

top_5_neighbourhoods <- total_auto_theft |>
  top_n(5, Total_Auto_Theft) |>
  pull(NEIGHBOURHOOD_140)

top_5_data <- train_data |>
  filter(NEIGHBOURHOOD_140 %in% top_5_neighbourhoods)

top_5_age_data <- top_5_data |>
  group_by(NEIGHBOURHOOD_140, REPORT_YEAR) |>
  summarise(
    Total_Population = sum(`0 to 14 years` + `15 to 24 years` + `25 to 54 years` + `55 to 64 years` + `65 years and over`, na.rm = TRUE),
    Percent_0_14 = sum(`0 to 14 years`, na.rm = TRUE) / Total_Population * 100,
    Percent_15_24 = sum(`15 to 24 years`, na.rm = TRUE) / Total_Population * 100,
    Percent_25_54 = sum(`25 to 54 years`, na.rm = TRUE) / Total_Population * 100,
    Percent_55_64 = sum(`55 to 64 years`, na.rm = TRUE) / Total_Population * 100,
    Percent_65_Plus = sum(`65 years and over`, na.rm = TRUE) / Total_Population * 100
  ) |>
  ungroup()

top_testing_summary <- top_5_data |>
  group_by(NEIGHBOURHOOD_140, REPORT_YEAR) |>
  summarise(
    Total_Auto_Theft = n(),
    Avg_Income = mean(`Average total income`, na.rm = TRUE)
  ) |>
  ungroup()

top_5_combined <- top_testing_summary |>
  left_join(top_5_age_data, by = c("NEIGHBOURHOOD_140", "REPORT_YEAR"))

age_distribution_long <- top_5_combined |>
  pivot_longer(
    cols = starts_with("Percent_"),
    names_to = "Age_Group",
    values_to = "Percentage"
  )

plot_ly(age_distribution_long, 
        x = ~NEIGHBOURHOOD_140, 
        y = ~Percentage, 
        color = ~Age_Group, 
        type = 'scatter', 
        mode = 'lines+markers', 
        text = ~paste0(Age_Group, ": ", round(Percentage, 1), "%"),
        hoverinfo = 'text',
        frame = ~REPORT_YEAR) |>
  layout(
    title = list(
      text = paste0(
        "Age Group Trends in Top 5 Auto Theft neighbourhoods by In ",
        "<br><sub><i>Animated line chart showing the % of each age group in Toronto in 2015 and 2020</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "neighbourhood"),
    yaxis = list(title = "Percentage"),
    margin = list(t = 100, l = 100),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "left",
        x = 0.1,
        y = -0.15,
        buttons = list(
          list(label = "Play", method = "animate", args = list(NULL)),
          list(label = "Pause", method = "animate", args = list(NULL, list(frame = list(duration = 0, redraw = FALSE))))
        )
      )
    )
  )
```


The plot shows that individuals aged 25–54 consistently make up the largest age group across the top five auto theft neighbourhoods in 2015, suggesting a relatively younger working-age population. Meanwhile, younger ages (0–14 years old) and older (55+ years) age groups represent smaller shares, with slight variation between neighbourhoods in both 2015 and 2020.


## Housing Plots {.tabset}
The following plots examine various housing types, like duplexes, high-rise buildings (over five storeys), low-rise buildings (under five storeys), semi-detached houses, single-detached houses, and row houses, and how they are distributed across Toronto neighbourhoods. The plots show how their prevalence may relate to auto theft incidents, and these comparisons helps highlight whether certain residential environments are more associated with higher rates of auto theft.

Each point represents one of Toronto’s 140 neighbourhoods, and for each neighbourhood, there are two points: one for data in 2015, and another point for data in 2020. 
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
plot_data <- train_data |>
  group_by(REPORT_YEAR, `Neighbourhood Name`) |>
  select(REPORT_YEAR, `Neighbourhood Name`, `Total_Auto_Thefts`, 
         `Single-detached house`, `Semi-detached house`, `Row house`,
         `Duplex Apt`, `Lowrise Apt`, `Highrise Apt`) |>
  pivot_longer(cols = 4:9, names_to = "Housing_Type", values_to = "Housing_Count")

housing_types <- unique(plot_data$Housing_Type)
```

### Single Detatched
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
single_detached_plot <- plot_ly(
  data = filter(plot_data, Housing_Type == "Single-detached house"),
  x = ~Housing_Count,
  y = ~Total_Auto_Thefts,
  color = ~`Neighbourhood Name`,
  colors = "Set1",
  text = ~paste(
    "neighbourhood:", `Neighbourhood Name`,
    "<br>Count:", Housing_Count,
    "<br>Thefts:", Total_Auto_Thefts,
    "<br>Year:", REPORT_YEAR
  ),
  type = "scatter",
  mode = "markers",
  marker = list(size = 10, opacity = 0.7)
) |>
  layout(
    title = list(
      text = paste0(
        "Single-Detached Homes vs Auto Thefts by neighbourhood",
        "<br><sub><i>Scatterplot showing the relationship between the number of single-detached housing units and auto thefts</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    margin = list(t = 100, l = 100),
    hoverlabel = list(align = "left")
  )
single_detached_plot
```

### Semi-Detatched 
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
semi_detached_plot <- plot_ly(
  data = filter(plot_data, Housing_Type == "Semi-detached house"),
  x = ~Housing_Count,
  y = ~Total_Auto_Thefts,
  color = ~`Neighbourhood Name`,
  colors = "Set2",
  text = ~paste(
    "neighbourhood:", `Neighbourhood Name`,
    "<br>Count:", Housing_Count,
    "<br>Thefts:", Total_Auto_Thefts,
    "<br>Year:", REPORT_YEAR
  ),
  type = "scatter",
  mode = "markers"
) |>
  layout(
    title = list(
      text = paste0(
        "Semi-Detached Homes vs Auto Thefts by neighbourhood",
        "<br><sub><i>Scatter plot showing the relationship between the number of semi-detached housing units and auto thefts</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    margin = list(t = 100, l = 100),
    hoverlabel = list(align = "left")
  )
semi_detached_plot
```


### Row House 
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
row_house_plot <- plot_ly(
  data = filter(plot_data, Housing_Type == "Row house"),
  x = ~Housing_Count,
  y = ~Total_Auto_Thefts,
  color = ~`Neighbourhood Name`,
  text = ~paste(
    "neighbourhood:", `Neighbourhood Name`,
    "<br>Count:", Housing_Count,
    "<br>Thefts:", Total_Auto_Thefts,
    "<br>Year:", REPORT_YEAR
  ),
  colors = "Set3",
  type = "scatter",
  mode = "markers"
) |>
  layout(
    title = list(
      text = paste0(
        "Row Houses vs Auto Thefts by neighbourhood",
        "<br><sub><i>Scatter plot showing the relationship between the number of row house units and auto thefts</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    margin = list(t = 100, l = 100),
    hoverlabel = list(align = "left")
  )
row_house_plot
```


### Duplex Apt  
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
duplex_plot <- plot_ly(
  data = filter(plot_data, Housing_Type == "Duplex Apt"),
  x = ~Housing_Count,
  y = ~Total_Auto_Thefts,
  color = ~`Neighbourhood Name`,
  text = ~paste(
    "neighbourhood:", `Neighbourhood Name`,
    "<br>Count:", Housing_Count,
    "<br>Thefts:", Total_Auto_Thefts,
    "<br>Year:", REPORT_YEAR
  ),
  colors = "Set2",
  type = "scatter",
  mode = "markers"
) |>
  layout(
    title = list(
      text = paste0(
        "Duplex Apartments vs Auto Thefts by neighbourhood",
        "<br><sub><i>Scatter plot showing the relationship between the number of duplex units and auto thefts</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    margin = list(t = 100, l = 100),
    hoverlabel = list(align = "left")
  )
duplex_plot
```

### Lowrise Apt  
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
lowrise_plot <- plot_ly(
  data = filter(plot_data, Housing_Type == "Lowrise Apt"),
  x = ~Housing_Count,
  y = ~Total_Auto_Thefts,
  color = ~`Neighbourhood Name`,
  text = ~paste(
    "neighbourhood:", `Neighbourhood Name`,
    "<br>Count:", Housing_Count,
    "<br>Thefts:", Total_Auto_Thefts,
    "<br>Year:", REPORT_YEAR
  ),
  colors = "Set3",
  type = "scatter",
  mode = "markers"
) |>
  layout(
    title = list(
      text = paste0(
        "Low-rise Apartments vs Auto Thefts by neighbourhood",
        "<br><sub><i>Scatter plot showing the relationship between the number of low-rises housing units and auto thefts</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    margin = list(t = 100, l = 100),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    hoverlabel = list(align = "left")
  )
lowrise_plot
```

### Highrise Apt  
```{r, echo=FALSE, error=FALSE, warning=FALSE, message=FALSE}
highrise_plot <- plot_ly(
  data = filter(plot_data, Housing_Type == "Highrise Apt"),
  x = ~Housing_Count,
  y = ~Total_Auto_Thefts,
  color = ~`Neighbourhood Name`,
  text = ~paste(
    "neighbourhood:", `Neighbourhood Name`,
    "<br>Count:", Housing_Count,
    "<br>Thefts:", Total_Auto_Thefts,
    "<br>Year:", REPORT_YEAR
  ),
  colors = "Set1",
  type = "scatter",
  mode = "markers"
) |>
  layout(
    title = list(
      text = paste0(
        "High-rises Apartments vs Auto Thefts by neighbourhood",
        "<br><sub><i>Scatter plot showing the relationship between the number of highrise units and auto thefts</i></sub>"
      ),
      x = 0.5
    ),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    margin = list(t = 100, l = 100),
    xaxis = list(title = "Number of Housing Units"),
    yaxis = list(title = "Number of Auto Thefts"),
    hoverlabel = list(align = "left")
  )
highrise_plot
```



##    
Most points cluster in the lower corner, suggesting moderate levels of both auto thefts and various housing types, with no strong or consistent relationship observed. However, these scatterplots are useful for identifying the overall distribution and spotting potential outliers among neighbourhoods.















